---
title: "sigident - Howto Microarray (short)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sigident-Howto_Microarray_short}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# An alternative workflow: four generic functions for signature identification 

In order to simplify the whole sigident-workflow described in the vignette [sigident - Howto Microarray](https://github.com/miracum/clearly-sigident/blob/master/vignettes/sigident-Howto_Microarray.Rmd), we wrapped all functions of the sigident R package framework into four big functions.

## Preprocessing

In order to use this R package and its functions, you need to prepare a merged gene expression dataset. The workflow to achieve this is presented in the following by making use of the R package `sigident.preproc`.

For a more detailed background and description of the following steps please view the `sigident.preproc` package's [vignette](https://github.com/miracum/clearly-sigident.preproc/blob/master/vignettes/sigident_preproc_Howto_GEO.Rmd).

## Initialization of important variables

```{r setup}
library(sigident)
library(sigident.preproc)
library(sigident.func)
library(knitr)
library(caret)

# initialize filePath:
filePath <- tempdir()

# define datadir
maindir <- "./geodata/"
datadir <- paste0(maindir, "data/")
dir.create(maindir)
dir.create(datadir)

# define plotdir
plotdir <- "./plots/"
dir.create(plotdir)

# define plotdir
csvdir <- "./csv/"
dir.create(csvdir)

# pathway
species <- "Hs"
OrgDb <- "org.Hs.eg.db"
organism <- "hsa"
pathwayid <- "hsa04151"

# diagnostig signature
seed <- 111
split <- 0.8
nfolds <- 10

# other variables
idtype = "affy"
fdr <- 0.05
```

## Define list that contains a representation of the studies metadata

```{r}
studiesinfo <- list(
  "GSE18842" = list(
    setid = 1,
    targetcolname = "source_name_ch1",
    targetlevelname = "Human Lung Tumor",
    controllevelname = "Human Lung Control"
    ),
  
  "GSE19804" = list(
    setid = 1,
    targetcolname = "source_name_ch1",
    targetlevelname = "frozen tissue of primary tumor",
    controllevelname = "frozen tissue of adjacent normal"
  ),
  
  "GSE19188" = list(
    setid = 1,
    targetcolname = "characteristics_ch1",
    controllevelname = "tissue type: healthy",
    targetlevelname = "tissue type: tumor",
    use_rawdata = TRUE
  )
)
```

## Load GEO datasets

All downloaded datasets will be assigned to the global environment.

```{r results='hide', message=FALSE, warning=FALSE, error=FALSE}
sigident.preproc::load_geo_data(
  studiesinfo = studiesinfo,
  datadir = datadir,
  plotdir = plotdir,
  idtype = idtype,
  viz_batch_boxp = F,
  viz_batch_gpca = F
) 
```

## Run `sigident.func::sigidentDEG`-function

```{r}
genes <- sigident.func::sigidentDEG(
  mergeset = mergeset,
  mergedset = mergedset,
  sample_metadata = sample_metadata,
  diagnosis = diagnosis,
  idtype = idtype,
  fdr = fdr,
  plotdir = plotdir,
  csvdir = csvdir
)
```

## Run `sigident.func::sigidentEnrichment`-function

```{r}
sigident.func::sigidentEnrichment(
  mergeset = mergeset,
  mergedset = mergedset,
  idtype = idtype,
  diagnosis = diagnosis,
  species = species,
  org_db = OrgDb,
  organism = organism,
  pathwayid = pathwayid,
  plotdir = plotdir,
  csvdir = csvdir
)
```
```{r}
filename <- paste0(plotdir, "/", pathwayid, ".png")
knitr::include_graphics(filename)
```
```{r}
filename <- paste0(plotdir, "/", pathwayid, ".pathview.png")
knitr::include_graphics(filename)
```

## Run `sigidentDiagnostic`-function

```{r}
# run only on first 1000 columns
colindices <- 1:1000
```


```{r}
diagnostic_models <- sigident::sigidentDiagnostic(
  mergeset = mergeset,
  diagnosis = diagnosis,
  seed = seed,
  nfolds = nfolds,
  split = split,
  plotdir = plotdir,
  colindices = colindices
)
```

```{r}
knitr::kable(
  sigident::compare_diagnostic_models(diagnostic_models)
)
```

```{r}
knitr::kable(
  sigident::get_diagnostic_lambda_values(diagnostic_models)
)
```

### Validation of the diagnostic signatures 

```{r}
validationstudylist <- list(
  "GSE30219" = list(
    setid = 1, 
    targetcolname = "source_name_ch1",
    controllevelname = "Non Tumoral Lung",
    targetlevelname = "Lung Tumour"
  ),
  "GSE33356" = list(
    setid = 1, 
    targetcolname = "characteristics_ch1",
    controllevelname = "tissue: paired normal adjacent",
    targetlevelname = "tissue: lung cancer"
  ),
  "GSE102287" = list(
    setid = 2, 
    targetcolname = "characteristics_ch1.6",
    controllevelname = "tumor_normal status: N",
    targetlevelname = "tumor_normal status: T"
  )
)
```

```{r}
validation_results <- sigident::validate_diagnostic_signatures(
  validationstudylist = validationstudylist,
  models = diagnostic_models,
  genes = genes,
  idtype = idtype,
  datadir = datadir,
  colindices = colindices
)
```

## Run `sigidentPrognostic`-Function 

### First, define meta information lists

```{r}
validationstudiesinfo <- list(
  "GSE30219" = list(
    setid = 1,
    timecol = "characteristics_ch1.9",
    status = list(
      statuscol = "characteristics_ch1.8",
      levels = list(alive = "status: ALIVE",
                    deceased = "status: DEAD",
                    na = "status: NTL")
    ),
    targetcolname = "source_name_ch1",
    controllevelname = "Non Tumoral Lung",
    targetlevelname = "Lung Tumour"),
  
  "GSE50081" = list(
    setid = 1,
    timecol = "characteristics_ch1.8",
    status = list(
      statuscol = "characteristics_ch1.9",
      levels = list(
        alive = "status: alive",
        deceased = "status: dead",
        na = NA)
    ),
    targetcolname = "characteristics_ch1.1",
    targetlevelname = paste(
      "histology: adenocarcinoma",
      "histology: adenosquamous carcinoma",
      "histology: large cell carcinoma",
      "histology: NSClarge cell carcinoma-mixed",
      "histology: NSCLC-favor adenocarcinoma",
      "histology: squamous cell carcinoma",
      "histology: squamous cell carcinoma X2",
      sep = "|||"
    ),
    controllevelname = NULL)
)
# if targetlevelname == NULL, the expression set will not be filtered further
# but instead, all samples will be used for calculations
```

```{r eval = TRUE}
discoverystudies_w_timedata <- list(
  "GSE19188" = list(
    setid = 1,
    timecol = "characteristics_ch1.2",
    status = list(
      statuscol = "characteristics_ch1.3",
      levels = list(alive = "status: alive",
                    deceased = "status: deceased",
                    na = "status: Not available")),
    targetcolname = "characteristics_ch1",
    controllevelname = "tissue type: healthy",
    targetlevelname = "tissue type: tumor",
    use_rawdata = TRUE)
)
```

### Then run the analysis 

```{r}
progn_results <- sigident::sigidentPrognostic(
  mergeset = mergeset,
  sample_metadata = sample_metadata,
  idtype = idtype,
  genes = genes,
  discoverystudies_w_timedata = discoverystudies_w_timedata,
  classifier_studies = c("GSE18842", "GSE19804"),
  validationstudiesinfo = validationstudiesinfo,
  datadir = datadir,
  plotdir = plotdir,
  csvdir = csvdir
)
```

```{r}
# create roc plot
filename <- paste0(plotdir, "GSE30219_Prognostic_Kaplan-Meier_Plot.png")
```

```{r}
knitr::include_graphics(filename)
```

```{r}
# create roc plot
filename <- paste0(plotdir, "GSE50081_Prognostic_Kaplan-Meier_Plot.png")
```

```{r}
knitr::include_graphics(filename)
```
